---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A1.** 32 departments, and Atmospheric and Oceanic Science has the most, with 93 classes

```{r}

climate_classes <- umd_courses |> filter (str_detect(title, "Climate") & seats > 0)

#sorting by department
climate_classes |> 
  group_by(department) |> 
  summarize(
    count_classes = n()
  ) |> 
  arrange(desc(count_classes))

```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? **A2.** Anthropology is not a department that I would've expected would offer the second most climate-related classes

**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3.**

Business and Management, School of Music, Information Studies, English, Teaching and Learning, Policy and Leadership had the most, and 134 departments have at least one syllabus in that term.

Business and management, economics, school of music, information studies, engineering (professional masters) were the top 5 among departments with at least one syllabus.

The difference between the two is the number of STEM classes, which I think are more pre-disposed to have posted a syllabus.

```{r}

#looking at departments in fall 2024
fall_2024_courses <- umd_courses |> filter (term == 202408)

fall_2024_courses |> 
  group_by(department) |> 
  summarize(
    count_classes = n()
  ) |> 
  arrange(desc(count_classes))
```

```{r}

#same but with syllabus
syllabus_fall_2024_courses <- umd_courses |> filter (term == 202408 & syllabus_count >= 1)

syllabus_fall_2024_courses |> 
  group_by(department) |> 
  summarize(
    count_classes = n()
  ) |> 
  arrange(desc(count_classes))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe.

Then create a new column for the month based on expenditure_date.

Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name).

Write a sentence describing the pattern of spending over time.

**A4.** The spending pattern starts relatively low, but generally increasing, and then jumps by a factor of 10 in the summer of 2022, reaching it's highest point of over \$3 million in October 2022.

```{r}

maryland_expenses <- maryland_expenses |> mutate(new_expenditure_date=mdy(expenditure_date))

```

```{r}
wes_moore <- maryland_expenses |> filter(str_detect(committee_name, "Wes For"))
```

```{r}

wes_moore |>
  mutate(month = floor_date(new_expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(month)
  
  
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).

**A5.** \$77723146 was spent in Maryland, and \$33164994 was spent in other states

```{r}

maryland_expenses |>
  mutate(
    spending_location = case_when(
      str_detect(address, " Maryland ") ~ "In-state",
      str_detect(address, " California ") ~ "Out of state",
      str_detect(address, " Washington ") ~ "Out of state",
      str_detect(address, " Louisiana ") ~ "Out of state",
      str_detect(address, " Florida ") ~ "Out of state",
      str_detect(address, " North Carolina ") ~ "Out of state",
      str_detect(address, " Massachusetts ") ~ "Out of state",
      str_detect(address, " West Virginia ") ~ "Out of state",
      str_detect(address, " Virginia ") ~ "Out of state",
      .default = NA
    )
  ) |> 
  group_by(spending_location) |> 
  summarise(
    total_amount = sum(amount)
  ) |> 
  arrange(desc(total_amount))

```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A6.** How much has spending increased from 2021-2022 across all payment methods? There is a massive increase across the board. (I tried to do just do spending without adding expense_method, but it wasn't working for some reason??)

```{r}
year_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))
```

```{r}

year_expenses |> mutate(year = floor_date(expenditure_date, "year")) |>
  group_by(year, expense_method) |>
  summarise(
    total_amount = sum(amount)
    ) |> 
  arrange(expense_method)


```

```{r}

#not sure why this doesnt work
maryland_expenses |> 
  group_by(expense_method)
  summarise(
    total_amount = sum(amount)
  )
```
