---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

payments <- read_csv(
  "data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", 
  guess_max = 369008
  ) |> 
  clean_names() |> 
  mutate(
    vendor_name = str_to_upper(vendor_name)
  )

# Load in the Maryland grants and loan data

md_grants_loans <- read_csv(
  "data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv"
  ) |> 
  clean_names() |> 
  mutate(
    grantor = str_to_upper(grantor), 
    grantee = str_to_upper(grantee), 
    description = str_to_upper(description), 
    category = str_to_upper(category)
    )

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

**A1:** 59 records

```{r}

pg_schools <- payments |> filter (
  str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOLS") | 
  str_detect(vendor_name, "PRINCE GEORGE'S COUNTY PUBLIC SCHOOLS") | 
  str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD") | 
  str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE'S COUNTY") | 
  str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE") | 
  str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") | 
  str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") |
  str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOLS") |
  str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOLS")
  )
  
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

**A2:** 2024 had the most amount, a billion dollars per year seems crazy to me for public school funding

```{r}

pg_schools |> 
  group_by(fiscal_year) |> 
  summarise(
    total_amount = sum(amount)
  ) |> 
  arrange(desc(total_amount))

```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

**A3.** First, you'll need to clean up the zip code row by removing all the ones that contain characters using case_when. Then you'll have to go through all the vendors, using case when to create a new column of cleaned vendors, changing the similar vendors to be fully matching.

A MORTON THOMAS & ASSOC INC and A MORTON THOMAS & ASSOCIATES INC

A&I INC and A&I INC.

AEROSOL MONITORING & ANALYSIS INC and AERSOL MONITORING & ANALYSIS INC

```{r}

umd_cap_projects <- payments |> filter (str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND"))

umd_cap_projects |> 
  group_by(vendor_name, ) |> 
  summarise(
    total_amount = sum(amount)
  ) |> 
  arrange(vendor_name)
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year.

Search the data to find grants that fit that description.

Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter.

Then calculate the total amount of money for each fiscal year.

Write a sentence summarizing how stem cell research funding has changed over time.

A4: Funding was high in 2009, then it's been fluctuating since then, hitting a low in 2021, and then jumping almost back up to the 2009 amount in 2023, perhaps representing renewed interest in stem cell research

```{r}

stem_cell_grants <- md_grants_loans |> 
  filter (str_detect(description, "STEM CELL"))

stem_cell_grants |> 
  group_by(fiscal_year) |> 
  summarise(
    total_amount = sum(amount)
  ) |> 
  arrange(fiscal_year)
```

Q5: Using the Maryland grants and loan data,

find records where the zip code is the one where the University of Maryland-College Park is located in

and produce a list of grantees from that zip code, along with the total amount they received.

Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

**A5:** I count 15 unique grantees. To clean the results so that there is only one record per grantee, I think you would have to use case_when again, going through each grantee and standardizing them one by one. It works in this case because there are so few records for this filter, but it would be a lot harder the more records there are.

I think the UMD data center grantee is interesting to me, simply because data centers are really under fire right now.

```{r}

md_grants_loans <- md_grants_loans |> 
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |> 
  mutate(zip5 = case_when(
    zip5 == "Vario" ~ NA,
    zip5 == "0" ~ NA,
    zip5 == "O8108" ~ "08108",
    str_detect(zip5, "UB7") ~ NA,
    .default = zip5
  ))



md_grants_loans |> 
  filter(str_detect(zip5, "20742")) |> 
  group_by(grantee) |> 
  summarise(
    total_amount = sum(amount)
  ) |> 
  arrange(grantee)


```
