---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

**A1:** It's from a website (https, no data/)

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:** There are 10489 records, which are all earthquakes (more specifically detectable tremors). We can look at date/time, location, magnitude, depth, type, and everything else we would need to identify each tremor.

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}

earthquakes |> 
  arrange(desc(depth))



```

**A3:** The deepest one was at a depth of 665.057 and had a magnitude of 4.20

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}

earthquakes |> 
  filter (mag > 6)




```

**A4:** There are 4 earthquakes with a magnitude greater than 6

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}

earthquakes |> 
  filter (mag > 6 & depth < 20)



```

**A5:** There is only one

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}

earthquakes |> 
  filter (mag > 6 | depth < 20)


```

**A6:** There are 8,247 earthquakes that fit this criteria

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}

earthquakes |> 
  filter (str_detect(place, "Alaska"))




```

**A7:** Look at the place column, then filter and use str_detect with "Alaska". There were 3,431 earthquakes.

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}

earthquakes |> 
  group_by(type) |> 
  summarise(
    count = n()
  ) |> 
  arrange(desc(count))



```

**A8:** The 2 most common types were earthquakes and quarry blasts.

------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}

earthquakes |> 
  group_by(type) |> 
  summarise(
    average = mean(depth)
  )


```

**A9:** The average depth for earthquakes in this data set is 21.9. It is odd that the average depth for 3 of the types is 0 (also, what is negative depth??)

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}

earthquakes <- earthquakes |> 
  mutate(hour = hour(time))



```

**A10:** Use mutate to create a new hour column and then use the hour function on the time column to pull the hour

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}
 earthquakes <- earthquakes |> 
  mutate(severity = case_when(
      mag > 3 ~ "serious",
      .default = "not serious"
    )
  )
  


```

**A11:** Use mutate and case_when to go through every row in mag and if the number is greater than 3, put serious in the new column, and if it's under 3, but not serious

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquakes |> 
  filter(severity == "serious") |> 
  group_by(hour) |> 
  summarise(
    count = n()
  ) |> 
  #arrange(desc(count)) #for most serious
  arrange(count) #for least serious


```

**A12**: The hour with the most serious earthquakes is a tie between 10:00 and 20:00, while the hour with the least amount of serious earthquakes is 6:00

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?


Which days have had the most serious earthquakes? September 12th
```{r}
earthquakes <- earthquakes |> 
  mutate (month_day = format(time,"%m-%d")) #found this online for creating a month + day column from ymd dates 


earthquakes |> 
  filter(severity == "serious") |> 
  group_by(month_day) |> 
  summarise(
    count = n()
  ) |> 
  arrange(desc(count))
  



```

```{r}

```

```{r}

```
